Splitting it up: the \pkg{spduration} split-population duration regression package

### 2015-10-27 manuscript.Rmd
new graphics, writing is next

We begin by loading the package and formatting the data appropriately. 
```{r, eval=FALSE}
library(spduration)
data(coup.data)
new.coup.data <- add_duration(coup.data, "coup", unitID="countryid", tID="year",
                              freq="year")
```

@@

```{r models, eval=FALSE}
weib_model <- spdur(
  duration ~ milreg + instab + rwar + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data=bscoup)

loglog_model <- spdur(
  duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
  atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
South.Am.+Central.Am.,
  data=bscoup, distr="loglog")
```

Using the \code{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values. One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects.  
\small
\begin{verbatim}
> AIC(weib.model)
[1] 1120.774
> AIC(loglog.model)
[1] 1111.908
> BIC(weib.model)
[1] 1209.577
> BIC(loglog.model)
[1] 1200.711
\end{verbatim}

\normalsize
The objects produced by \texttt{spdur} can also be used with functions in the \texttt{xtable} package to produce \LaTeX\ code for tables of results. The \texttt{xtable} function used on an \texttt{spdur} object will produce code such as that used to create Table \ref{loglog_table}. 
\small
\begin{verbatim}
> xtable(loglog.model,caption="Coup model with log-logistic hazard")
\end{verbatim}

\begin{table}[ht]
\centering
\caption{Coup model with log-logistic hazard} \label{loglog_table}
\begin{tabular}{rlrrr}
  \hline
 & Parameter & Estimate & StdErr & p \\ 
  \hline
1 & (Dur. Intercept) & 3.29 & 0.14 & 0.00 \\ 
  2 & Military.Regime & -1.51 & 0.16 & 0.00 \\ 
  3 & Instability & -0.21 & 0.04 & 0.00 \\ 
  4 & Recent.War & -0.20 & 0.28 & 0.48 \\ 
  5 & Regional.Conflict & 2.64 & 2.56 & 0.30 \\ 
  6 & log(alpha) & -0.53 & 0.07 & 0.00 \\ 
  7 & (Risk Intercept) & -13.10 & 10.03 & 0.19 \\ 
  8 & Coup.Risk & 7.32 & 3.23 & 0.02 \\ 
  9 & GDP.cap. & 3.11 & 1.82 & 0.09 \\ 
  10 & Military.Regime.1 & 25.41 & 0.01 & 0.00 \\ 
  11 & Recent.War.1 & -4.34 & 2.70 & 0.11 \\ 
  12 & Regional.Conflict.1 & -39.74 & 23.96 & 0.10 \\ 
  13 & South.Am. & -3.72 & 3.19 & 0.24 \\ 
  14 & Central.Am. & -6.08 & 2.45 & 0.01 \\ 
   \hline
\end{tabular}
\end{table}

\normalsize
The table shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation. Interestingly, the coup risk variable has a rather large and statistically significant coefficient in the risk equation.

The package also includes a function, \texttt{plot.hazard}, that can be used with an \texttt{spdur} object to plot the estimated hazard rate. \texttt{plot.hazard} plots the conditional hazard, which is the probability of survival conditional on the covariates in the risk and duration equations, and conditional on survival up to time $t$. The function calculates the estimated hazard rate as well as 90\% confidence intervals, which are produced by simulating values from the estimated sampling distributions of the model parameters. By default the function uses the mean values of the covariates during the simulations, but users can choose specific covariate values by entering them as vectors in the arguments \texttt{xvals} and \texttt{zvals}, which correspond to the covariates in the duration and risk equations, respectively. The command below creates Figure \ref{hazard}.
\begin{verbatim}
> plot.hazard(loglog.model)
\end{verbatim}

\begin{figure}[htbp!]
\centering
\includegraphics[width=4in]{graphics/loglog_hazard.pdf}
\caption{Conditional hazard rate from log-logistic model.} \label{hazard}
\end{figure}

Importantly, the package also includes functions to evaluate model predictions. The generic function \texttt{predict} can be used on an object of class \texttt{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period. In addition, the generic \texttt{plot} function can be used on \texttt{spdur} objects to produce a separation plot [@greenhill2011separation], which is a graphical display for evaluating model predictions. The code below produces Figure \ref{insamp}.
\small
\begin{verbatim}
> dev.new(width=9,height=3)
> par(mfrow=c(2,1),mar=c(2,2,2,2))
> plot(weib.model,endSpellOnly=F)
> plot(loglog.model,endSpellOnly=F)
\end{verbatim}

\normalsize
The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every observation, not only those at the end of a spell, is used in the plot. By default the \texttt{plot} function will calculate the conditional hazard for each observation. The separation plot sorts observations from left to right according to the predicted probability assigned by the model (higher values to the right), and shows each event/failure as red line, with non-events shown in beige. This makes it easy to see whether the model is assigning high probabilities of failure to actual cases of failure, and low probabilities to non-failures. 
\begin{figure}[htbp!]
\centering
\includegraphics[width=5in]{graphics/in-sample.pdf}
\caption{{\em In-sample separation plots for Weibull and log-logistic models.}} \label{insamp}
\end{figure}

Finally, we demonstrate use of the \texttt{spdurCrisp} function for evaluating out-of-sample predictions. We begin by splitting the data into training and test sets, using the last year of data (1999) as the test set and omitting incomplete observations from the test set. 
\small
\begin{verbatim}
> coup.train<-new.coup.data[coup.data$year<1999,]
> coup.test<-new.coup.data[new.coup.data$year==1999,]
> coup.test<-na.omit(coup.test)
\end{verbatim}

\normalsize
The \texttt{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set. 
\small
\begin{verbatim}
> weib.model2<-spdurCrisp(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+ 		atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
+ 		South.Am.+Central.Am.,
+ 		last='end.spell',train=coup.train,test=coup.test,pred=coup.test)
> weib.preds<-as.numeric(weib.model2$test.p)
>
> loglog.model2<-spdurCrisp(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+ 		atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
+ 		South.Am.,
+ 		last='end.spell',train=coup.train,test=coup.test,pred=coup.test,distr="loglog")
> loglog.preds<-as.numeric(loglog.model2$test.p)
\end{verbatim}

\normalsize
The \texttt{train},\texttt{test}, and \texttt{pred} arguments are used to specify data frames for the training set, test set, and forecasting set. We do not wish to generate forecasts here, so we simply specify the same data frame for the test and forecasting set. Here we use \texttt{spdurCrisp} to calculate the conditional risk, which is the default calculation for this function and is the probability that an observation is in the at-risk population. The test set predictions are stored in the model object as \texttt{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \texttt{separationplot} package. 
\small
\begin{verbatim}
> dev.new(width=9,height=3)
> par(mfrow=c(2,1),mar=c(2,2,2,2))
> separationplot(weib.preds,coup.test$Coup,newplot=F,show.expected=T,lwd1=5,lwd2=2)
> separationplot(loglog.preds,coup.test$Coup,newplot=F,show.expected=T,lwd1=5,lwd2=2)
\end{verbatim}

![](out-of-sample.pdf)

\normalsize
\begin{figure}[htbp!]
\centering
\includegraphics[width=5in]{out-of-sample.pdf}
\caption{{\em Out-of-sample separation plots for Weibull and log-logistic models. }} \label{outsamp}
\end{figure}

@@

```{r rates, include = FALSE}
library(foreign)
library(magrittr)
library(ggplot2)
library(xtable)
library(scales)

fl2003 <- read.dta("data/repdata.dta")
rates <- data.frame(
  data = "Fearon and Laitin 2003",
  rate = mean(fl2003$onset > 0)
)

load("data/irc_data_mod.rda")
rates <- rbind(rates, data.frame(
  data = "Beger et al 2015",
  rate = mean(irc.data$irr.t)
))

belkin <- read.dta("data/BelkinSchoferTable4.dta")
belkin$coup <- ifelse(belkin$coup=="yes", 1, 0)
rates <- rbind(rates, data.frame(
  data = "Belkin & Schofer 2003",
  rate = mean(belkin$coup)
))

rates$rate <- rates$rate * 1000

# Convert data to factor ordered by rate, so plot will sort by rate
rates$data <- with(rates, factor(data, levels = data[order(rate)]))

#xtable(rates, digits = 2)

p <- ggplot(rates, aes(x = data, y = rate)) + 
  geom_bar(stat = "identity", fill = muted("blue"), width = 0.6) +
  coord_flip() +
  theme_minimal() +
  labs(y = "Positives per 1,000", x = "")

ggsave(plot = p, file = "graphics/rates.pdf", height = 1.3, width = 5, scale = 1)
```

>>>>

We begin by loading the package and formatting the data appropriately. 
```{r data}
library("spduration")
#data(bscoup)
library("foreign")
bscoup <- read.dta("data/BelkinSchoferTable4.dta")
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                            freq="year")
```

@@

```{r models, include = FALSE}
weib_model <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data=bscoup, silent = TRUE)

loglog_model <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data=bscoup, distr="loglog", silent = TRUE)
```

Using the \code{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values. One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects. 
```{r fit-stats}
AIC(weib_model)
AIC(loglog_model)
BIC(weib_model)
BIC(loglog_model)
```

The objects produced by \texttt{spdur} can also be used with functions in the \texttt{xtable} package to produce \LaTeX\ code for tables of results. The \texttt{xtable} function used on an \texttt{spdur} object will produce code such as that used to create Table \ref{loglog_table}.

```{r summary-table, include=FALSE}
library("xtable")
tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard",
              label="loglog_table")
print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
```

\begin{table}[ht]
\centering
\caption{Coup model with log-logistic hazard} 
\label{loglog_table}
\begin{tabular}{lrrr}
  \hline
Parameter & Estimate & StdErr & p \\ 
  \hline
(Dur. Intercept) & 3.30 & 0.14 & 0.00 \\ 
  milreg & -1.53 & 0.15 & 0.00 \\ 
  instab & -0.22 & 0.03 & 0.00 \\ 
  regconf & 3.04 & 2.22 & 0.17 \\ 
  log(alpha) & -0.52 & 0.07 & 0.00 \\ 
  (Risk Intercept) & -14.61 & 9.02 & 0.11 \\ 
  couprisk & 9.11 & 4.37 & 0.04 \\ 
  wealth & 3.51 & 1.75 & 0.05 \\ 
  milreg.1 & 21.76 & 0.24 & 0.00 \\ 
  rwar & -6.37 & 5.38 & 0.24 \\ 
  regconf.1 & -20.22 & 54.04 & 0.71 \\ 
  samerica & -3.31 & 3.28 & 0.31 \\ 
  camerica & -6.46 & 2.57 & 0.01 \\ 
   \hline
\end{tabular}
\end{table}

The table shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation. Interestingly, the coup risk variable has a rather large and statistically significant coefficient in the risk equation.

The package also includes a function, \texttt{plot.hazard}, that can be used with an \texttt{spdur} object to plot the estimated hazard rate. \texttt{plot.hazard} plots the conditional hazard, which is the probability of survival conditional on the covariates in the risk and duration equations, and conditional on survival up to time $t$. The function calculates the estimated hazard rate as well as 90\% confidence intervals, which are produced by simulating values from the estimated sampling distributions of the model parameters. By default the function uses the mean values of the covariates during the simulations, but users can choose specific covariate values by entering them as vectors in the arguments \texttt{xvals} and \texttt{zvals}, which correspond to the covariates in the duration and risk equations, respectively. The command below creates Figure \ref{plot-hazard2}.

```{r hazard-ex-1, eval = FALSE}
plot(loglog_model, type="hazard", main = "A")
```

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/hazard-ex.pdf}
\end{center}
\end{figure}

```{r hazard-ex-2, eval = FALSE}
plot(loglog_model, type="hazard", 
     xvals = c(1, 1, 10, 0.05), 
     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
     main = "B")
```

Importantly, the package also includes functions to evaluate model predictions. The generic function \texttt{predict} can be used on an object of class \texttt{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period. In addition, the generic \texttt{plot} function can be used on \texttt{spdur} objects to produce a separation plot [@greenhill2011separation], which is a graphical display for evaluating model predictions. The code below produces Figure \ref{insamp}.

```{r, eval = FALSE}
plot(weib_model)
plot(loglog_model)
```

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/sepplots.pdf}
\end{center}
\end{figure}

The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every observation, not only those at the end of a spell, is used in the plot. By default the \texttt{plot} function will calculate the conditional hazard for each observation. The separation plot sorts observations from left to right according to the predicted probability assigned by the model (higher values to the right), and shows each event/failure as red line, with non-events shown in beige. This makes it easy to see whether the model is assigning high probabilities of failure to actual cases of failure, and low probabilities to non-failures.

Finally, we demonstrate how to evaluate a model's out of sample predictions. We begin by splitting the data into training and test sets, using the last year of data (1999) as the test set and omitting incomplete observations from the test set. 
```{r}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
                          freq="year")
```

```{r}
coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                           freq="year")
coup_test  <- coup_test[coup_test$year >= 1996, ]
```

\normalsize
The \texttt{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set.

```{r}
weib_model2   <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = coup_train, silent = TRUE)

loglog_model2 <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = coup_train, distr="loglog", silent = TRUE) 
```

Foo

```{r}
weib2_test_p   <- predict(weib_model2, newdata = coup_test)
loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
```

\normalsize
The \texttt{train},\texttt{test}, and \texttt{pred} arguments are used to specify data frames for the training set, test set, and forecasting set. We do not wish to generate forecasts here, so we simply specify the same data frame for the test and forecasting set. Here we use \texttt{spdurCrisp} to calculate the conditional risk, which is the default calculation for this function and is the probability that an observation is in the at-risk population. The test set predictions are stored in the model object as \texttt{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \texttt{separationplot} package.

```{r oos-sepplots}
library("separationplot")

obs_y <- coup_test$coup[complete.cases(coup_test)]

par(mfrow=c(2,1),mar=c(2,2,2,2))
separationplot(weib2_test_p,   obs_y)
separationplot(loglog2_test_p, obs_y)
```

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
\end{center}
\end{figure}

```{r}
glm_model2    <- glm(
  coup ~ milreg + instab + regconf + couprisk + wealth + rwar + regconf + 
    samerica + camerica,
  data = coup_train, family = "binomial"
)

c(Weibull = AIC(weib_model2), 
  Loglog = AIC(loglog_model2), 
  Logistic = AIC(glm_model2))
```

@@

### 2015-10-27 manuscript.tex
new graphics, writing is next

\begin{CodeChunk}
\begin{CodeInput}
library(spduration)
data(coup.data)
new.coup.data <- add_duration(coup.data, "coup", unitID="countryid", tID="year",
                              freq="year")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
weib_model <- spdur(
  duration ~ milreg + instab + rwar + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data=bscoup)

loglog_model <- spdur(
  duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
  atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
South.Am.+Central.Am.,
  data=bscoup, distr="loglog")
\end{CodeInput}
\end{CodeChunk}

Using the \code{summary} function on either model object will produce
standard output showing the model formula, estimates for the duration
and risk equations, and test statistics with \(p\)-values. One may also
use the \code{AIC} and \code{BIC} function to calculate the information
criterion statistics for \code{spdur} objects.\\\small

\begin{verbatim}
> AIC(weib.model)
[1] 1120.774
> AIC(loglog.model)
[1] 1111.908
> BIC(weib.model)
[1] 1209.577
> BIC(loglog.model)
[1] 1200.711
\end{verbatim}

\normalsize
The objects produced by \texttt{spdur} can also be used with functions
in the \texttt{xtable} package to produce \LaTeX~code for tables of
results. The \texttt{xtable} function used on an \texttt{spdur} object
will produce code such as that used to create Table \ref{loglog_table}.
\small

\begin{verbatim}
> xtable(loglog.model,caption="Coup model with log-logistic hazard")
\end{verbatim}

\begin{table}[ht]
\centering
\caption{Coup model with log-logistic hazard} \label{loglog_table}
\begin{tabular}{rlrrr}
  \hline
 & Parameter & Estimate & StdErr & p \\ 
  \hline
1 & (Dur. Intercept) & 3.29 & 0.14 & 0.00 \\ 
  2 & Military.Regime & -1.51 & 0.16 & 0.00 \\ 
  3 & Instability & -0.21 & 0.04 & 0.00 \\ 
  4 & Recent.War & -0.20 & 0.28 & 0.48 \\ 
  5 & Regional.Conflict & 2.64 & 2.56 & 0.30 \\ 
  6 & log(alpha) & -0.53 & 0.07 & 0.00 \\ 
  7 & (Risk Intercept) & -13.10 & 10.03 & 0.19 \\ 
  8 & Coup.Risk & 7.32 & 3.23 & 0.02 \\ 
  9 & GDP.cap. & 3.11 & 1.82 & 0.09 \\ 
  10 & Military.Regime.1 & 25.41 & 0.01 & 0.00 \\ 
  11 & Recent.War.1 & -4.34 & 2.70 & 0.11 \\ 
  12 & Regional.Conflict.1 & -39.74 & 23.96 & 0.10 \\ 
  13 & South.Am. & -3.72 & 3.19 & 0.24 \\ 
  14 & Central.Am. & -6.08 & 2.45 & 0.01 \\ 
   \hline
\end{tabular}
\end{table}

\normalsize
The table shows estimates from the duration equation, beginning with the
intercept, and then estimates from the risk equation. Interestingly, the
coup risk variable has a rather large and statistically significant
coefficient in the risk equation.

The package also includes a function, \texttt{plot.hazard}, that can be
used with an \texttt{spdur} object to plot the estimated hazard rate.
\texttt{plot.hazard} plots the conditional hazard, which is the
probability of survival conditional on the covariates in the risk and
duration equations, and conditional on survival up to time \(t\). The
function calculates the estimated hazard rate as well as 90\% confidence
intervals, which are produced by simulating values from the estimated
sampling distributions of the model parameters. By default the function
uses the mean values of the covariates during the simulations, but users
can choose specific covariate values by entering them as vectors in the
arguments \texttt{xvals} and \texttt{zvals}, which correspond to the
covariates in the duration and risk equations, respectively. The command
below creates Figure \ref{hazard}.

\begin{verbatim}
> plot.hazard(loglog.model)
\end{verbatim}

\begin{figure}[htbp!]
\centering
\includegraphics[width=4in]{graphics/loglog_hazard.pdf}
\caption{Conditional hazard rate from log-logistic model.} \label{hazard}
\end{figure}

Importantly, the package also includes functions to evaluate model
predictions. The generic function \texttt{predict} can be used on an
object of class \texttt{spdur} to generate several kinds of predictions,
including the probability that an observation is ``at-risk'' and the
probability of failure for a given time period. In addition, the generic
\texttt{plot} function can be used on \texttt{spdur} objects to produce
a separation plot (Greenhill, Ward, and Sacks 2011), which is a
graphical display for evaluating model predictions. The code below
produces Figure \ref{insamp}. \small

\begin{verbatim}
> dev.new(width=9,height=3)
> par(mfrow=c(2,1),mar=c(2,2,2,2))
> plot(weib.model,endSpellOnly=F)
> plot(loglog.model,endSpellOnly=F)
\end{verbatim}

\normalsize
The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every
observation, not only those at the end of a spell, is used in the plot.
By default the \texttt{plot} function will calculate the conditional
hazard for each observation. The separation plot sorts observations from
left to right according to the predicted probability assigned by the
model (higher values to the right), and shows each event/failure as red
line, with non-events shown in beige. This makes it easy to see whether
the model is assigning high probabilities of failure to actual cases of
failure, and low probabilities to non-failures.

\begin{figure}[htbp!]
\centering
\includegraphics[width=5in]{graphics/in-sample.pdf}
\caption{{\em In-sample separation plots for Weibull and log-logistic models.}} \label{insamp}
\end{figure}

Finally, we demonstrate use of the \texttt{spdurCrisp} function for
evaluating out-of-sample predictions. We begin by splitting the data
into training and test sets, using the last year of data (1999) as the
test set and omitting incomplete observations from the test set. \small

\begin{verbatim}
> coup.train<-new.coup.data[coup.data$year<1999,]
> coup.test<-new.coup.data[new.coup.data$year==1999,]
> coup.test<-na.omit(coup.test)
\end{verbatim}

\normalsize
The \texttt{spdurCrisp} function can then be used to fit a model on the
training set and generate predictions for the test set. \small

\begin{verbatim}
> weib.model2<-spdurCrisp(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+       atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
+       South.Am.+Central.Am.,
+       last='end.spell',train=coup.train,test=coup.test,pred=coup.test)
> weib.preds<-as.numeric(weib.model2$test.p)
>
> loglog.model2<-spdurCrisp(duration~Military.Regime+Instability+Recent.War+Regional.Conflict,
+       atrisk~Coup.Risk+GDP.cap.+Military.Regime+Recent.War+Regional.Conflict+
+       South.Am.,
+       last='end.spell',train=coup.train,test=coup.test,pred=coup.test,distr="loglog")
> loglog.preds<-as.numeric(loglog.model2$test.p)
\end{verbatim}

\normalsize
The \texttt{train},\texttt{test}, and \texttt{pred} arguments are used
to specify data frames for the training set, test set, and forecasting
set. We do not wish to generate forecasts here, so we simply specify the
same data frame for the test and forecasting set. Here we use
\texttt{spdurCrisp} to calculate the conditional risk, which is the
default calculation for this function and is the probability that an
observation is in the at-risk population. The test set predictions are
stored in the model object as \texttt{test.p}, and we extract them to
produce the plot shown below in Figure \ref{outsamp}, which was created
using the \texttt{separationplot} package. \small

\begin{verbatim}
> dev.new(width=9,height=3)
> par(mfrow=c(2,1),mar=c(2,2,2,2))
> separationplot(weib.preds,coup.test$Coup,newplot=F,show.expected=T,lwd1=5,lwd2=2)
> separationplot(loglog.preds,coup.test$Coup,newplot=F,show.expected=T,lwd1=5,lwd2=2)
\end{verbatim}

\includegraphics{out-of-sample.pdf}

\normalsize

\begin{figure}[htbp!]
\centering
\includegraphics[width=5in]{out-of-sample.pdf}
\caption{{\em Out-of-sample separation plots for Weibull and log-logistic models. }} \label{outsamp}
\end{figure}

>>>>

\begin{CodeChunk}
\begin{CodeInput}
library("spduration")
#data(bscoup)
library("foreign")
bscoup <- read.dta("data/BelkinSchoferTable4.dta")
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                            freq="year")
\end{CodeInput}
\begin{CodeOutput}
Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
yyyy-06-30
\end{CodeOutput}
\end{CodeChunk}

@@

Using the \code{summary} function on either model object will produce
standard output showing the model formula, estimates for the duration
and risk equations, and test statistics with \(p\)-values. One may also
use the \code{AIC} and \code{BIC} function to calculate the information
criterion statistics for \code{spdur} objects.

\begin{CodeChunk}
\begin{CodeInput}
AIC(weib_model)
\end{CodeInput}
\begin{CodeOutput}
[1] 1207.992
\end{CodeOutput}
\begin{CodeInput}
AIC(loglog_model)
\end{CodeInput}
\begin{CodeOutput}
[1] 1110.829
\end{CodeOutput}
\begin{CodeInput}
BIC(weib_model)
\end{CodeInput}
\begin{CodeOutput}
[1] 1290.452
\end{CodeOutput}
\begin{CodeInput}
BIC(loglog_model)
\end{CodeInput}
\begin{CodeOutput}
[1] 1193.289
\end{CodeOutput}
\end{CodeChunk}

The objects produced by \texttt{spdur} can also be used with functions
in the \texttt{xtable} package to produce \LaTeX~code for tables of
results. The \texttt{xtable} function used on an \texttt{spdur} object
will produce code such as that used to create Table \ref{loglog_table}.

\begin{table}[ht]
\centering
\caption{Coup model with log-logistic hazard} 
\label{loglog_table}
\begin{tabular}{lrrr}
  \hline
Parameter & Estimate & StdErr & p \\ 
  \hline
(Dur. Intercept) & 3.30 & 0.14 & 0.00 \\ 
  milreg & -1.53 & 0.15 & 0.00 \\ 
  instab & -0.22 & 0.03 & 0.00 \\ 
  regconf & 3.04 & 2.22 & 0.17 \\ 
  log(alpha) & -0.52 & 0.07 & 0.00 \\ 
  (Risk Intercept) & -14.61 & 9.02 & 0.11 \\ 
  couprisk & 9.11 & 4.37 & 0.04 \\ 
  wealth & 3.51 & 1.75 & 0.05 \\ 
  milreg.1 & 21.76 & 0.24 & 0.00 \\ 
  rwar & -6.37 & 5.38 & 0.24 \\ 
  regconf.1 & -20.22 & 54.04 & 0.71 \\ 
  samerica & -3.31 & 3.28 & 0.31 \\ 
  camerica & -6.46 & 2.57 & 0.01 \\ 
   \hline
\end{tabular}
\end{table}

The table shows estimates from the duration equation, beginning with the
intercept, and then estimates from the risk equation. Interestingly, the
coup risk variable has a rather large and statistically significant
coefficient in the risk equation.

The package also includes a function, \texttt{plot.hazard}, that can be
used with an \texttt{spdur} object to plot the estimated hazard rate.
\texttt{plot.hazard} plots the conditional hazard, which is the
probability of survival conditional on the covariates in the risk and
duration equations, and conditional on survival up to time \(t\). The
function calculates the estimated hazard rate as well as 90\% confidence
intervals, which are produced by simulating values from the estimated
sampling distributions of the model parameters. By default the function
uses the mean values of the covariates during the simulations, but users
can choose specific covariate values by entering them as vectors in the
arguments \texttt{xvals} and \texttt{zvals}, which correspond to the
covariates in the duration and risk equations, respectively. The command
below creates Figure \ref{plot-hazard2}.

\begin{CodeChunk}
\begin{CodeInput}
plot(loglog_model, type="hazard", main = "A")
\end{CodeInput}
\end{CodeChunk}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/hazard-ex.pdf}
\end{center}
\end{figure}

\begin{CodeChunk}
\begin{CodeInput}
plot(loglog_model, type="hazard", 
     xvals = c(1, 1, 10, 0.05), 
     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
     main = "B")
\end{CodeInput}
\end{CodeChunk}

Importantly, the package also includes functions to evaluate model
predictions. The generic function \texttt{predict} can be used on an
object of class \texttt{spdur} to generate several kinds of predictions,
including the probability that an observation is ``at-risk'' and the
probability of failure for a given time period. In addition, the generic
\texttt{plot} function can be used on \texttt{spdur} objects to produce
a separation plot (Greenhill, Ward, and Sacks 2011), which is a
graphical display for evaluating model predictions. The code below
produces Figure \ref{insamp}.

\begin{CodeChunk}
\begin{CodeInput}
plot(weib_model)
plot(loglog_model)
\end{CodeInput}
\end{CodeChunk}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/sepplots.pdf}
\end{center}
\end{figure}

The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every
observation, not only those at the end of a spell, is used in the plot.
By default the \texttt{plot} function will calculate the conditional
hazard for each observation. The separation plot sorts observations from
left to right according to the predicted probability assigned by the
model (higher values to the right), and shows each event/failure as red
line, with non-events shown in beige. This makes it easy to see whether
the model is assigning high probabilities of failure to actual cases of
failure, and low probabilities to non-failures.

Finally, we demonstrate how to evaluate a model's out of sample
predictions. We begin by splitting the data into training and test sets,
using the last year of data (1999) as the test set and omitting
incomplete observations from the test set.

\begin{CodeChunk}
\begin{CodeInput}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
                          freq="year")
\end{CodeInput}
\begin{CodeOutput}
Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
yyyy-06-30
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                           freq="year")
\end{CodeInput}
\begin{CodeOutput}
Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
yyyy-06-30
\end{CodeOutput}
\begin{CodeInput}
coup_test  <- coup_test[coup_test$year >= 1996, ]
\end{CodeInput}
\end{CodeChunk}

\normalsize
The \texttt{spdurCrisp} function can then be used to fit a model on the
training set and generate predictions for the test set.

\begin{CodeChunk}
\begin{CodeInput}
weib_model2   <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = coup_train, silent = TRUE)

loglog_model2 <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = coup_train, distr="loglog", silent = TRUE) 
\end{CodeInput}
\end{CodeChunk}

Foo

\begin{CodeChunk}
\begin{CodeInput}
weib2_test_p   <- predict(weib_model2, newdata = coup_test)
loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
\end{CodeInput}
\end{CodeChunk}

\normalsize
The \texttt{train},\texttt{test}, and \texttt{pred} arguments are used
to specify data frames for the training set, test set, and forecasting
set. We do not wish to generate forecasts here, so we simply specify the
same data frame for the test and forecasting set. Here we use
\texttt{spdurCrisp} to calculate the conditional risk, which is the
default calculation for this function and is the probability that an
observation is in the at-risk population. The test set predictions are
stored in the model object as \texttt{test.p}, and we extract them to
produce the plot shown below in Figure \ref{outsamp}, which was created
using the \texttt{separationplot} package.

\textbackslash{}begin\{CodeChunk\} \textbackslash{}begin\{CodeInput\}
library(``separationplot'')

obs\_y \textless{}- coup\_test\$coup{[}complete.cases(coup\_test){]}

par(mfrow=c(2,1),mar=c(2,2,2,2)) separationplot(weib2\_test\_p, obs\_y)
separationplot(loglog2\_test\_p, obs\_y)
\textbackslash{}end\{CodeInput\} \textbackslash{}end\{CodeChunk\}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
\end{center}
\end{figure}

\begin{CodeChunk}
\begin{CodeInput}
glm_model2    <- glm(
  coup ~ milreg + instab + regconf + couprisk + wealth + rwar + regconf + 
    samerica + camerica,
  data = coup_train, family = "binomial"
)

c(Weibull = AIC(weib_model2), 
  Loglog = AIC(loglog_model2), 
  Logistic = AIC(glm_model2))
\end{CodeInput}
\begin{CodeOutput}
  Weibull    Loglog  Logistic 
 990.5818  985.2633 1171.3556 
\end{CodeOutput}
\end{CodeChunk}
